Statistical Inference for inter-arrival times of extreme events in bursty time series

### 2020-03-26 article_springer/article_springer.tex
Überarbeitung, Finale Änderung bzgl Reviewer 2

In the following we denote with \(x_L \in [-\infty, +\infty)\) and
\(x_R \in (-\infty, +\infty]\)) the left and right endpoint of the
distribution of \(J\). We assume that the \(k\)-th magnitude \(J_k\)
occurs at time \(T_k = W_1 + \ldots + W_k\). Based on an MRP, we define
the Continuous Time Random Exceedance model (CTRE) as follows:

@@

In this article, we restrict ourselves to the uncoupled case, where
\(W\) and \(J\) are independent. Then the two sequences
\(X(\ell, n)_{n \in \mathbb N}\) and \(T(\ell, n)_{n \in \mathbb N}\)
are independent as well. To see why, note that \(X(\ell)\) is, in
distribution, simply equal to \(J - \ell | J > \ell\), independent of
any waiting time \(W_k\).\\
We assume for the rest of the article, that the magnitudes
\((J_i)_{i \in \mathbb{N}}\) belong to the the max-domain of attraction
of some non-degenerate distribution. This means there exist \(a_n>0\)
and \(d_n \in \mathbb{R}\) such that
\[ a_n^{-1}(J_1 \vee \ldots \vee J_n -d_n) \Rightarrow A \text{ as } n \rightarrow \infty.\]
Hence, the distribution of \(A\) is a generalized extreme value
distribution (GEV) which distribution function is given by
\[ F(x;\xi) = \begin{cases}\exp(-(1+\xi x)^{-1/\xi}) & \xi\neq0 \\ \exp(-\exp(-x)) & \xi = 0\end{cases}.\]
The GEV is subdivided into the Gumbel (\(\xi=0\)), the Weibull
(\(\xi<0\)) and the Fr'echet (\(\xi>0\)) family of distributions.

@@

\emph{Proof of Theorem:} We interpret the threshold crossing time
\(T(\ell)\) as the hitting time of the underlying CTRM (Continuous Time
Random Maxima) or ``max-renewal process'', and then utilize a result by
Meerschaert and Stoev (2008). The running maximum process is defined as
\[
M(c) := J_1 \vee \ldots \vee J_{\lfloor c \rfloor},
\] and since we assume that the \(J_k\) have a continuous distribution,
there exist norming functions \(a(c)\) and \(d(c)\) such that \[
\mathbf P\left[ \frac{M(c) - d(c)}{a(c)} \le \ell^* \right] 
\longrightarrow F(\ell^*), \quad c \to \infty
\] where \(F\) is a generalized extreme value distribution, and
\(\ell^*\) is any value from the support of \(F\). The CTRM process is
then defined via \[
V(t) = M(N(t)), \quad t \ge 0
\] where \(N(t)\) is the renewal process associated with the waiting
times \(W_k\): \[
N(t) = \max\{n: W_1 + \ldots + W_n \le t\}.
\] Now a key observation is that \[
T(\ell) = \inf\{t: V(t) > \ell\}, 
\] and that

@@

The classical POT approach is based on the fact that exceedances above a
high threshold are asymptotically GPD distributed, hence a GPD
distribution is fitted to the exceedances above a high enough threshold.
Due to the Theorem in the last Section we know that for high thresholds
the times between exceedances are asymptotically Mittag-Leffler
distributed in case of heavy-tailed waiting times between the
observations. Hence, before we talk about inference for the exceedances
in the next section, we discuss inference for Mittag-Leffler
distributions.

Historically, the first method proposed for the estimation of the
Mittag-Leffler distribution parameters was the fractional moment
estimator by Kozubowski (2001). Unlike the first moments, the fractional
moments of order \(p\) for \(p<\beta\) exist and are tractable. One
drawback of this method is that constant priors for the tail parameter
are needed for the calculation of the estimates. Cahoy et al. (2010)
proposed a moment estimator of the log-transformed data, which does not
require any prior. Furthermore, they performed simulation studies
illustrating that the log-moment outperforms the fractional moment
estimator with respect to bias and root mean squared error (RMSE).

Due to the form of the density function of a Mittag-Leffler distribution
there exists no closed form for the Maximum Likelihood estimator (MLE).
In the R Package \texttt{MittagLeffleR} (Gill and Straka, 2017), Maximum
Likelihood estimation is implemented via numerical optimization. The MLE
slightly outperforms the log-moment estimator regarding bias and RMSE
for big enough sample sizes, but is extremely computationally intensive.
Figure \ref{fig:MSE} shows that both estimators perform well, even for
small sample sizes.

@@

\caption{\label{fig:MSE} MSE for the estimation of tail (left column) and scale (right column) parameters via log-moment estimator and MLE of a Mittag-Leffler sample with varying sample size n=25, 100, 200, varying tails on the x-axis and fixed scale equal to one.}\label{fig:MSE}
\end{figure}

Since the Mittag-Leffler distribution is heavy-tailed, many researchers
would intuitively give the highest importance to the tail behaviour of
the distribution. Of course, one can also use established tail exponent
estimators for the estimation of the parameter \(\beta\) such as the
Hill estimator (Hill, 1975) based on the \(r+1\) upper order statistics
\begin{align}\label{eq:Hill}
H_{r,n}=\left[ \frac{1}{r} \sum_{i=0}^{r} \log \frac{X_{(i)}}{X_{(r+1)}}\right]^{-1}, 
\end{align} where \(X_{(1)} \geq X_{(2)} \geq \ldots \geq X_{(n)}\)
denote the order statistics in decreasing order of a sequence
\(X_1,\ldots,X_n\). However, these methods are statistically less
efficient since they only use a portion of the information contained in
the data, as mentioned by Kozubowski (2001). Moreover, the Hill
estimator requires a tuning parameter, denoted with \(r\) in
\eqref{eq:Hill}. Additionaly, the estimator only performs reliably well
for distributions close to Pareto (see Resnick, 1997). Figure
\ref{fig:Hillplots} shows Hill plots for Mittag-Leffler simulated data,
with varying sample sizes and tails. To deduce the correct tail
parameter estimates from these plots is virtually impossible. To
estimate the tail parameter of a Mittag-Leffler distribution with the
Hill estimator, the sample size has to be much larger and one has to
choose a large tuning parameter \(r\). In case that only events above a
high threshold are recorded, even \(n=1000\) events are unrealistic.
Furthermore the Hill estimator of course completely fails if the
inter-arrival times are exponentially distributed and not heavy-tailed.
We will come back to the problems with the Hill estimtator in Section
\ref{Simulationstudy}.

@@

\caption{\label{fig:Hillplots} Hillplots for simulated Mittag-Leffler data with true tail 0.8 and sample size 200 (left) and 1000 (right), with number of upper order statics r on which the Hill estimator is based on the x-axis. }\label{fig:Hillplots}
\end{figure}

Since the exponential distribution is nested in the Mittag-Leffler
family of distributions, a Likelihood-ratio Test (LRT) seems to be an
appropriate way to choose between a model with exponential and
Mittag-Leffler inter-exceedance times. Although the two models are
nested, the asymptotic distribution is not \(\chi^2_1\)-distributed, and
Wilk's Theorem does not hold: under \(H_0\), the parameter \(\beta\) of
Mittag-Leffler distribution is equal to \(1\), and hence lies on the
boundary of the parameter space \((0,1]\). Instead, a valid approach is
a bootstrapped Likelihood-ratio test (see e.g.~Davison et al., 1997).
Figure \ref{Fig:LRT} displays the (simulated) power for the bootstrapped
LRT for Mittag-Leffler distributions with varying tail parameters based
on 1000 simulation runs. As expected, the power decreases for tail
parameters close to one, since the Mittag-Leffler distribution converges
as \(\beta \uparrow 1\) to an exponential distribution; it becomes hard
to differentiate a Mittag-Lefller distribution from an exponential.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Extract the \(K\) largest order statistics (i.e.~the \(K\) largest
  values, where e.g.~\(K = 300\)) \(X_{(1)}, \ldots, X_{(K)}\) together
  with their timestamps \(T_{(1)}, \ldots, T_{(K)}\).
\item
  Choose e.g.~\(K_0 = 5\), and for each \(k\) ranging from \(K_0 +1\) to
  \(K + 1\):

  \begin{enumerate}
  \def\labelenumii{\alph{enumii})}
  \tightlist
  \item
    extract the set \(\mathcal T_k\) of exceedance times of the
    threshold \(X_{(k)}\)
  \item
    fit a Mittag-Leffler distribution to \(\mathcal T_k\), resulting in
    the parameter estimates \(\hat\beta_k\) and \(\hat \sigma_k\).
  \end{enumerate}
\item
  Plot \(k\) vs.~\(\hat \beta_k\). To the right, as
  \(k \downarrow K_0\), the asymptotics are off and bias is high. To the
  left (high threshold), data are scarce and variance is high. Choose a
  region of \emph{stability} in the middle for a parameter estimate
  \(\hat \beta\).
\item
  Plot \(k\) vs.~\(k^{1/\hat \beta} \hat \sigma_k\). Where the plot
  stabilizes as \(k \downarrow K_0\), define the value
  \(\hat \sigma_0\).
\end{enumerate}

@@

To clarify Step 4: Recall that by the theorem,
\(\sigma_k / b(1/p_{\ell})\) is expected to stabilize around a constant
as \(k \downarrow K_0\). Since \(b\) is regularly varying with parameter
\(1/\beta\), we have
\(b(1/p_{\ell}) = p_{\ell}^{-1/\beta} / L(1/p_{\ell})\) for some slowly
varying function \(L\). Approximating
\(p_{\ell} \approx \hat p_{\ell} := k / n\), we have
\[\text{const} \approx \sigma_k / b(1/p_{\ell}) = \sigma_k \times p_{\ell}^{1/\beta} L(1/p_{\ell}) \approx \sigma_k \times k^{1/\beta} n^{-1/\beta} L(n/k)\]
Assuming that the variation of \(L(n/k)\) is minor, we can hence see
that \(\sigma_k k^{1/\beta}\) stabilizes.

@@

To test our inference method via stability plots, we have simulated
\(m=100\) times \(n=10000\) independent waiting time and magnitude pairs
\((W_k, J_k)\) for waiting times that follow

@@

\begin{itemize}
\tightlist
\item
  rows correspond to cases (i), (ii) and (iii), and
\item
  columns correspond to estimators (log-moment estimator,
  maximum-likelihood, and Hill estimator).
\end{itemize}

We plot the tail parameter estimates \(\hat \beta(k)\) against \(k\) for
each of the \(m=100\) simulation runs. Thin grey lines represent
individual simulation runs, and the thicker black line is their mean.
Recall that \(k\) is the index of the order statistics of \(J_k\) at
which the threshold \(\ell\) is placed. Note that the Hill estimator is
based on the waiting times, not on the inter-exceedance times, meaning
that the estimation is based on more data. If one were to base the
Hill-estimation only on the inter-exceedance times, the sample size
would be too small and it would be impossible to deduce an estimate for
the tail parameter from the plots (compare Figure \ref{fig:Hillplots}).

In the right-hand Hill-estimator column, the \(k\) on the \(x\)-axis
corresponds to \(r\) from \eqref{eq:Hill}. The Hill estimator appears to
be nicely unbiased on Pareto data, but compared to the other two
estimators, it has a higher variance, as can be seen from the wider
spread of the simulated curves. This high variance means that that a
single estimate is less reliable compared to the other two estimators.
This effect is amplified if one only has a thresholded, and thus smaller
dataset available, to the extent that the Hill estimates become severely
flawed (see Section \ref{sec:ML}).

Moreover, the bottom in \ref{Fig:TailSimu} shows clearly that the
log-moment and maximum-likelihod estimators generalize to the
exponential case (\(\beta = 1\)), whereas the Hill-estimator does not.

Finally, the performance of Log-Moment and Maximum-Likelihood estimators
for the scale parameter is shown in figure \ref{Fig:ScaleSimu}. Since
there is no Hill-estimate of a scale parameter, there are only two
columns. Both estimators show good performance, with a slight advantage
for the Maximum-Likelihood estimator. This advantage is paid for by a
higher computational cost.

@@

\caption{\label{Fig:TailSimu} Stability Plots for m=100 simulation runs for stable distributed waiting times with a tail parameter of 0.8 (top row), Pareto distributed waiting times with tail parameter 0.8 (middle row) and exponentially distributed waiting times (lower row). Left column: log-moment estimator, middle column: MLE, right column: Hill estimator. The grey thin lines are the stability plots for the different simulation runs and the dark lines are their means. The red dotted line shows the true tail parameter. }\label{fig:TailSimuplots}
\end{figure}

@@

{\centering \includegraphics[width=0.9\linewidth]{article_springer_files/figure-latex/ScaleSimuplots-1}

@@

\caption{\label{Fig:ScaleSimu} Stability Plots for m=100 simulation runs for stable distributed waiting times with a tail parameter of 0.8 (top row), Pareto distributed waiting times with tail parameter 0.8 (middle row) and exponentially distributed waiting times (lower row). Left column: log-moment estimator, right column: MLE. The grey thin lines are the stability plots for the different simulation runs and the dark lines are their means. The red dotted line shows the true tail parameter.}\label{fig:ScaleSimuplots}
\end{figure}

@@

Before we apply the POT approach described in Section 5 to the solar
flare data, we first have to check if all model assumptions are
fulfilled. The CTRE model is based on three main assumptions, which are
repeated below. For each assumption, we suggest one means of checking if
it holds:

\begin{description}
\item[i.i.d.:]
After removing the ``noise observations'' below the smallest threshold
\(\ell_0\), the pair sequence \((T(\ell_0, i), X(\ell_0,i))\) is i.i.d.
An indication if this is true is given by an auto-correlation plot for
the logarithms (to ensure finite moments) of the two time series.
\item[Uncoupled:]
Each \(T(\ell, i)\) is independent of each \(X(\ell, i)\). We propose an
empirical copula plot to check for any dependence.
\item[\({\rm ML}(\beta, \sigma)\) distribution of \(T(\ell, i)\):]
Apply a cutoff at the lowest threshold \(\ell_0\), extract the threshold
crossing times, and create a QQ Plot for the Mittag-Leffler
distribution. Use a log-moment estimate of the tail parameter for the
theoretical / population quantiles of the plot.
\end{description}

Figures \ref{fig:flare-diagnostics-1}, \ref{fig:flare-diagnostics-2} and
\ref{fig:flare-diagnostics-3} show the diagnostic plots for a minimum
threshold chosen at the 200th order statistic. There is some residual
autocorrelation for the sequence of threshold exceedance times that is
not accounted for by the CTRE model.

\begin{figure}
\includegraphics[width=\textwidth]{article_springer_files/figure-latex/flare-diagnostics-1-1} \caption{\label{fig:flare-diagnostics-1} Diagnostic plots for the solar flare data: auto-correlation function.}\label{fig:flare-diagnostics-1}
\end{figure}

@@

Figure \ref{fig:flares} shows the stability plots for the solar flare
data, on the left for the tail parameter and on the right for the scale
parameter. The dark grey ranges correspond to 95\% confidence intervals,
which are derived from the asymptotic normality of the log-moments
estimators (Cahoy, 2013) and the \(\delta\)-method (Gill and Straka,
2017); dashed lines show the deduced true values of \(\beta\)
resp.~\(\sigma_0\). The stability plot for the tail stabilizes nicely
around 0.85 (dashed line), while the scale parameter stabilizes less
obviously near \(3 \times 10^7\) (dashed line). The growth of the scale
parameter for lower threshold appears to be closer to linear in
\(p_{\ell}\), rather than proportional to \(p_{\ell}^{1/0.85}\) as
suggested by the Mittag-Leffler fits. The reason for this is likely that
the overall goodness of fit as compared to an exponential distribution
is improved due to the peaked shape of the Mittag-Leffler distribution
near \(0\), rather than its tail behaviour at \(\infty\). The reported
fit should hence come with the caveat that a Mittag-Leffler distribution
models exceedance times well only up to certain time-scales. More
research is needed into the modelling of scale transitions, where
inter-exceedance times appear to have different power laws across
different time scales.

\begin{figure}
\includegraphics[width=\textwidth]{article_springer_files/figure-latex/solar-flare-tail-scale-1} \caption{\label{fig:flares} Stability plots for the tail and scale parameter of the Mittag-Leffler distribution of the Solar Flare dataset. Dotted horizontal lines are at $\beta = 0.85$ and $\sigma_0 = 3 \times 10^7$ seconds $\approx 0.95$ years.}\label{fig:solar-flare-tail-scale}
\end{figure}

The fit with a Mittag-Leffler distribution (\(\beta = 0.85\)) is good
(see Figure \ref{fig:flare-diagnostics-3}), though there are signs that
the power-law tail tapers off for very large inter-threshold crossing
times. There is no apparent dependence between threshold exceedance
times and event magnitudes seen in the copula plot (see Figure
\ref{fig:flare-diagnostics-2}). We also counduct a bootstrapped LRT for
the null hypothesis of exponentially distributed inter-arrival times and
received a \(p\)-value of \(p<0.01\).

@@

We have extended the POT (Peaks-over-Threshold) model, a mainstay of
extreme value theory, to ``bursty'' time series, which have been studied
intensively in statistical physics. Burstiness is characterized by
power-law waiting times between events, and we have shown that the
Mittag-Leffler distribution arises naturally as a scaling limit for the
inter-exceedance times of high thresholds. Moreover, we have derived the
following non-linear scaling behaviour:
\(\sigma \sim p_{\ell}^{-1/\beta}\), where \(\sigma\) is the scale
parameter of the distribution of threshold exceedance times,
\(p_{\ell}\) is the fraction of magnitudes above the threshold, and
\(\beta\) the exponent of the power law. This ``anomalous'' scaling
behaviour in the bursty setting entails two phenomena:

>>>>

In the following we denote with \(x_L \in [-\infty, +\infty)\) and
\(x_R \in (-\infty, +\infty]\) the left and right endpoint of the
distribution of \(J\). We assume that the \(k\)-th magnitude \(J_k\)
occurs at time \(T_k = W_1 + \ldots + W_k\). Based on an MRP, we define
the Continuous Time Random Exceedance model (CTRE) as follows:

@@

In this article, we restrict ourselves to the uncoupled case, where
\(W\) and \(J\) are independent. Then the two sequences
\(X(\ell, n)_{n \in \mathbb N}\) and \(T(\ell, n)_{n \in \mathbb N}\)
are independent as well. To see why, note that \(X(\ell)\) is, in
distribution, simply equal to \(J - \ell | J > \ell\), independent of
any waiting time \(W_k\).\\
We assume for the rest of the article, that the magnitudes
\((J_i)_{i \in \mathbb{N}}\) belong to the the max-domain of attraction
of some non-degenerate distribution. This means there exist \(a_n>0\)
and \(d_n \in \mathbb{R}\) such that \begin{align} \label{assumptionJs}
a_n^{-1}(J_1 \vee \ldots \vee J_n -d_n) \Rightarrow A \text{ as } n \rightarrow \infty.
\end{align} Hence, the distribution of \(A\) is a generalized extreme
value distribution (GEV) which distribution function is given by
\[ F(x;\xi) = \begin{cases}\exp(-(1+\xi x)^{-1/\xi}) & \xi\neq0 \\ \exp(-\exp(-x)) & \xi = 0\end{cases}.\]
The GEV is subdivided into the Gumbel (\(\xi=0\)), the Weibull
(\(\xi<0\)) and the Fréchet (\(\xi>0\)) family of distributions.

@@

\emph{Proof of Theorem:} We interpret the threshold crossing time
\(T(\ell)\) as the hitting time of the underlying CTRM (Continuous Time
Random Maxima) or ``max-renewal process'', and then utilize a result by
Meerschaert and Stoev (2008). The running maximum process is defined as
\[
M(c) := J_1 \vee \ldots \vee J_{\lfloor c \rfloor},
\] and since we assumed that the \(J_k\) our out of the domain of
attraction of a max-stable distribution \(F\) (see
\eqref{assumptionJs}), there exist norming functions \(a(c)\) and
\(d(c)\) such that \[
\mathbf P\left[ \frac{M(c) - d(c)}{a(c)} \le \ell^* \right] 
\longrightarrow F(\ell^*), \quad c \to \infty
\] where \(F\) is a generalized extreme value distribution, and
\(\ell^*\) is any value from the support of \(F\). The CTRM process is
then defined via \[
V(t) = M(N(t)), \quad t \ge 0
\] where \(N(t)\) is the renewal process associated with the waiting
times \(W_k\): \[
N(t) = \max\{n: W_1 + \ldots + W_n \le t\}.
\] Now a key observation is that \[
T(\ell) = \inf\{t: V(t) > \ell\}, 
\] and that

@@

The classical POT approach is based on the fact that exceedances above a
high threshold are asymptotically GPD distributed, hence a GPD
distribution is fitted to the exceedances above a high enough threshold.
This approach was well-established a long time ago and is standard in
extreme value theory nowadays. Hence we just refer here to the according
literature (Smith (1984), Leadbetter (1991), Davison and Smith (1990)),
a nice introduction can be found in Beirlant et al. (2006) Ch.5.3,
Embrechts et al. (2013) Ch. 6.5. or Coles (2001) Ch.4. Hence, we focus
in this article mainly on the inter-exceedance times. Due to the Theorem
in the last Section we know that for high thresholds the times between
exceedances are asymptotically Mittag-Leffler distributed in case of
heavy-tailed waiting times between the observations. In this article we
propose to proceed with inter-exceedance times in a similar fashion as
with the magnitudes in the classical POT approach, more precisely to fit
a Mittag-Leffler distribution via stability plots to the
inter-exceedance times. Hence, before we talk about inference for the
inter-exceedance times in the next section, we first discuss inference
for Mittag-Leffler distributions.

Historically, the first method proposed for the estimation of the
Mittag-Leffler distribution parameters was the fractional moment
estimator by Kozubowski (2001). Unlike the first moments, the fractional
moments of order \(p\) for \(p<\beta\) exist and are tractable. One
drawback of this method is that constant priors for the tail parameter
are needed for the calculation of the estimates. Cahoy et al. (2010)
proposed a moment estimator of the log-transformed data, which does not
require any prior. Furthermore, they performed simulation studies
illustrating that the log-Moment outperforms the fractional moment
estimator with respect to bias and root mean squared error (RMSE).

Due to the form of the density function of a Mittag-Leffler distribution
there exists no closed form for the Maximum Likelihood estimator (MLE).
In the R Package \texttt{MittagLeffleR} (Gill and Straka, 2017), Maximum
Likelihood estimation is implemented via numerical optimization. The MLE
slightly outperforms the log-Moment estimator regarding bias and RMSE
for big enough sample sizes, but is extremely computationally intensive.
Figure \ref{fig:MSE} shows that both estimators perform well, even for
small sample sizes.

@@

\caption{\label{fig:MSE} MSE for the estimation of tail (left column) and scale (right column) parameters via log-Moment estimator and MLE of a Mittag-Leffler sample with varying sample size n=25, 100, 200, varying tails on the x-axis and fixed scale equal to one, based on 5000 simulation runs.}\label{fig:MSE}
\end{figure}

Since the Mittag-Leffler distribution is heavy-tailed, many researchers
would intuitively give the highest importance to the tail behaviour of
the distribution. Of course, one can also use established tail exponent
estimators for the estimation of the parameter \(\beta\) such as the
Hill estimator (Hill, 1975) based on the \(r+1\) upper order statistics
\begin{align}\label{eq:Hill}
H_{r,n}=\left[ \frac{1}{r} \sum_{i=0}^{r} \log \frac{X_{(i)}}{X_{(r+1)}}\right]^{-1}, 
\end{align} where \(X_{(1)} \geq X_{(2)} \geq \ldots \geq X_{(n)}\)
denote the order statistics in decreasing order of a sequence
\(X_1,\ldots,X_n\). The Hill estimator only performs reliably well for
distributions close to Pareto (see Resnick, 1997). It heavily depends on
the choice of \(r\) in \eqref{eq:Hill}, hence on the number of order
statistics entering into \eqref{eq:Hill}. Figure \ref{fig:Hillplots}
shows Hill plots for Mittag-Leffler simulated data, with varying sample
sizes and tails. To deduce the correct tail parameter estimates from
these plots is virtually impossible. To estimate the tail parameter of a
Mittag-Leffler distribution with the Hill estimator, the sample size has
to be much larger and one has to choose a large tuning parameter \(r\).
In case that only events above a high threshold are recorded,
\(n=10000\) events are unrealistic. There is a vast literature on
bias-reduced versions of the Hill estimator and on the optimal choice of
\(r\), mainly based on imposed second-order regular variation conditions
(see e.g.~Beirlant et al., 2006, p. Ch.4.4 and 4.7 and references in
it). Also, there are lot of other tail-index estimators not taken under
consideration in our article, since this would go beyond the scope of
our article. However, we know from the Theorem in Section
\ref{sec:scaling} that the inter-exceedance times are asymptotically
Mittag-Leffler distributed. So all these general tail (and scale)
estimation methods are statistically less efficient since they only use
a portion of the information contained in the data, as mentioned by
Kozubowski (2001). Furthermore, the Hill estimator of course completely
fails, if the inter-arrival times are exponentially distributed and not
heavy-tailed. Hence, we propose to use the log-Moment estimator or the
MLE for Mittag-Leffler distributions.

@@

\caption{\label{fig:Hillplots} Hillplots for simulated Mittag-Leffler data with true tail 0.8 and sample size 200 (left) and 10000 (right), with number of upper order statics r on which the Hill estimator is based on the x-axis. }\label{fig:Hillplots}
\end{figure}

Since the exponential distribution is nested in the Mittag-Leffler
family of distributions, a Likelihood-ratio Test (LRT) seems to be an
appropriate way to choose between a model with exponential and
Mittag-Leffler inter-exceedance times. Although the two models are
nested, the asymptotic distribution is not \(\chi^2_1\)-distributed, and
Wilk's Theorem does not hold: under \(H_0\), the parameter \(\beta\) of
Mittag-Leffler distribution is equal to \(1\), and hence lies on the
boundary of the parameter space \((0,1]\). Instead, a valid approach is
a bootstrapped Likelihood-ratio test (see e.g.~Davison et al., 1997).
Figure \ref{Fig:LRT} displays the (simulated) power for the bootstrapped
LRT for Mittag-Leffler distributions with varying tail parameters based
on 1000 simulation runs. As expected, the power decreases for tail
parameters close to one, since the Mittag-Leffler distribution converges
as \(\beta \uparrow 1\) to an exponential distribution; it becomes hard
to differentiate a Mittag-Leffler distribution from an exponential.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Extract the \(K\) largest order statistics (i.e.~the \(K\) largest
  values, where e.g.~\(K = 300\)) \(X_{(1)}, \ldots, X_{(K)}\) together
  with their timestamps \(T_{(1)}, \ldots, T_{(K-1)}\).
\item
  Choose a minimum number of exceedances \(K_0\leq K\),
  e.g.~\(K_0 = 5\), and for each \(k\) ranging from \(K_0\) to \(K\):

  \begin{enumerate}
  \def\labelenumii{\alph{enumii})}
  \tightlist
  \item
    extract the set \(\mathcal T_k\) of exceedance times between the
    magnitudes exceeding the threshold \(X_{(k)}\)
  \item
    fit a Mittag-Leffler distribution to \(\mathcal T_k\), resulting in
    the parameter estimates \(\hat\beta_k\) and \(\hat \sigma_k\).
  \end{enumerate}
\item
  Plot \(k\) vs.~\(\hat \beta_k\). To the right, as
  \(k \downarrow K_0\), the asymptotics are off and bias is high. To the
  left (high threshold), data are scarce and variance is high. Choose a
  region of \emph{stability} in the middle for a parameter estimate
  \(\hat \beta\).
\item
  Plot \(k\) vs.~\(k^{1/\hat \beta} \hat \sigma_k\). Again, choose a
  region of \emph{stability} as \(k\) gets smaller and define
  \(\hat \sigma_0\).
\end{enumerate}

@@

To clarify Step 4: Recall that by the theorem,
\(\sigma_k / b(1/p_{\ell})\) is expected to stabilize around a constant
as \(k\) gets smaller. Since \(b\) is regularly varying with parameter
\(1/\beta\), we have
\(b(1/p_{\ell}) = p_{\ell}^{-1/\beta} / L(1/p_{\ell})\) for some slowly
varying function \(L\). Approximating \(p_{\ell}\) by
\(\hat p_{\ell} := k / n\), we have
\[\text{const} \approx \sigma_k / b(1/p_{\ell}) = \sigma_k \times p_{\ell}^{1/\beta} L(1/p_{\ell}) \approx \sigma_k \times k^{1/\beta} n^{-1/\beta} L(n/k)\]
Assuming that the variation of \(L(n/k)\) is minor, we can hence see
that \(\sigma_k k^{1/\beta}\) stabilizes.

\begin{description}
\item[\textbf{Remark}:]
We approximated \(p_{\ell}\), the probability that an event is larger
than \(l\), by its relative frequency. One could also approximate this
tail probability by the GPD distribution fitted to the exceedances.
\end{description}

@@

To test our inference method via stability plots, we have simulated
\(m=100\) datasets with \(n=10000\) independent waiting time and
magnitude pairs \((W_k, J_k)\) for waiting times that follow

@@

\begin{itemize}
\tightlist
\item
  rows correspond to cases (i), (ii) and (iii), and
\item
  columns correspond to estimators (log-Moment estimator,
  Maximum-Likelihood).
\end{itemize}

We plot the tail parameter estimates \(\hat \beta(k)\) against \(k\) for
each of the \(m=100\) simulation runs. Thin grey lines represent
individual simulation runs, and the thicker black line is their mean.
Recall that \(k\) is the index of the order statistics of \(J_k\) at
which the threshold \(\ell\) is placed.

The performance of log-Moment and Maximum-Likelihood estimators for the
tail resp. the scale parameter is shown in Figure \ref{Fig:TailSimu}
resp. \ref{Fig:ScaleSimu}. Both estimators show good performance, with a
slight advantage for the Maximum-Likelihood estimator. This advantage is
paid for by a higher computational cost. Moreover, the bottom row in
Figure \ref{Fig:TailSimu} shows clearly that the log-Moment and
Maximum-Likelihood estimators generalize to the exponential case
(\(\beta = 1\)) whereas a general tail estimator like the Hill-estimator
would not.

@@

\caption{\label{Fig:TailSimu} Stability Plots for m=100 simulation runs for stable distributed waiting times with a tail parameter of 0.8 (top row), Pareto distributed waiting times with tail parameter 0.8 (middle row) and exponentially distributed waiting times (lower row). Left column: log-Moment estimator, right column: MLE. The grey thin lines are the stability plots for the different simulation runs and the dark lines are their means. The red dotted line shows the true tail parameter.}\label{fig:TailSimuplots}
\end{figure}

@@

{\centering \includegraphics[width=1\linewidth]{article_springer_files/figure-latex/ScaleSimuplots-1}

@@

\caption{\label{Fig:ScaleSimu} Stability Plots for m=100 simulation runs for stable distributed waiting times with a tail parameter of 0.8 (top row), Pareto distributed waiting times with tail parameter 0.8 (middle row) and exponentially distributed waiting times (lower row). Left column: log-Moment estimator, right column: MLE. The grey thin lines are the stability plots for the different simulation runs and the dark lines are their means. The red dotted line shows the true scale parameter.}\label{fig:ScaleSimuplots}
\end{figure}

@@

Before we apply the approach described in Section 5 to the solar flare
data, we first have to check if all model assumptions are fulfilled. The
CTRE model is based on three main assumptions, which are repeated below.
For each assumption, we suggest one means of checking if it holds:

\begin{description}
\item[i.i.d.:]
After removing the ``noise observations'' below the smallest threshold
\(\ell_0\), the pair sequence \((T(\ell_0, i), X(\ell_0,i))\) is i.i.d.
An indication if this is true is given by an auto-correlation plot.
Since we are expecting the inter-exceedance times to be Mittag-Leffler
distributed and hence to have infinite mean but finite log-moments, we
first take the logarithms of the times.
\item[Uncoupled:]
Each \(T(\ell, i)\) is independent of each \(X(\ell, i)\). We propose an
empirical copula plot to check for any dependence.
\item[\({\rm ML}(\beta, \sigma)\) distribution of \(T(\ell, i)\):]
Apply a cutoff at the lowest threshold \(\ell_0\), extract the threshold
crossing times, and create a QQ Plot for the Mittag-Leffler
distribution. Use a Log-moment estimate of the tail parameter for the
theoretical / population quantiles of the plot.
\item[\textbf{Remark:}]
The ACF plots of course can just give an indication whether there are
dependencies, since they actually just measure linear dependencies.
Furthermore, if one calculates the ACF for the logarithmic
inter-exceedance times, the ACF indicates on the original scale a
multiplicative dependence.
\end{description}

Figures \ref{fig:flare-diagnostics-1}, \ref{fig:flare-diagnostics-2} and
\ref{fig:flare-diagnostics-3} show the diagnostic plots for a minimum
threshold chosen at the 100th order statistic. There is some residual
autocorrelation for the sequence of threshold exceedance times that is
not accounted for by the CTRE model.

\begin{figure}
\includegraphics[width=\textwidth]{article_springer_files/figure-latex/flare-diagnostics-1-1} \caption{\label{fig:flare-diagnostics-1} Diagnostic plots for the solar flare data based on the 100 upper order statistics: auto-correlation and cross-correalation function.}\label{fig:flare-diagnostics-1}
\end{figure}

@@

Figure \ref{fig:flares} shows the stability plots for the solar flare
data, on the left for the tail parameter and on the right for the scale
parameter. The dark grey ranges correspond to 95\% confidence intervals,
which are derived from the asymptotic normality of the Log-moment
estimators (Cahoy, 2013) and the \(\delta\)-method (Gill and Straka,
2017); dashed lines show the deduced true values of \(\beta\)
resp.~\(\sigma_0\). The stability plot for the tail stabilizes nicely
around 0.9 (dashed line), while the scale parameter stabilizes less
obviously near \(3 \times 10^7\) (dashed line). The growth of the scale
parameter for lower threshold appears to be closer to linear in
\(p_{\ell}\), rather than proportional to \(p_{\ell}^{1/0.9}\) as
suggested by the Mittag-Leffler fits. The reason for this is likely that
the overall goodness of fit as compared to an exponential distribution
is improved due to the peaked shape of the Mittag-Leffler distribution
near \(0\), rather than its tail behaviour at \(\infty\). The reported
fit should hence come with the caveat that a Mittag-Leffler distribution
models exceedance times well only up to certain time-scales. More
research is needed into the modelling of scale transitions, where
inter-exceedance times appear to have different power laws across
different time scales.

\begin{figure}
\includegraphics[width=\textwidth]{article_springer_files/figure-latex/solar-flare-tail-scale-1} \caption{\label{fig:flares} Stability plots for the tail and scale parameter of the Mittag-Leffler distribution of the Solar Flare dataset. Dotted horizontal lines are at $\beta = 0.9$ and $\sigma_0 = 3 \times 10^7$ seconds $\approx 0.95$ years.}\label{fig:solar-flare-tail-scale}
\end{figure}

The fit with a Mittag-Leffler distribution (\(\beta = 0.9\)) seems to be
good (see Figure \ref{fig:flare-diagnostics-3}), though there are signs
that the power-law tail tapers off for very large inter-threshold
crossing times. There is no apparent dependence between threshold
exceedance times and event magnitudes seen in the copula plot (see
Figure \ref{fig:flare-diagnostics-2}). We also conduct a bootstrapped
LRT for the null hypothesis of exponentially distributed inter-arrival
times and received a \(p\)-value of \(p<0.01\).

@@

We have extended the POT (Peaks-over-Threshold) model, a mainstay of
extreme value theory, to ``bursty'' time series, which have been studied
intensively in statistical physics. Burstiness is characterized by
power-law waiting times between events, and we have shown that the
Mittag-Leffler distribution arises naturally as a scaling limit for the
inter-exceedance times of high thresholds. Moreover, we have derived the
following non-linear scaling behaviour:
\(\sigma \sim p_{\ell}^{-1/\beta}\), where \(\sigma\) is the scale
parameter of the distribution of threshold exceedance times,
\(p_{\ell}\) is the fraction of magnitudes above the threshold \(\ell\),
and \(\beta\) the exponent of the power law. This ``anomalous'' scaling
behaviour in the bursty setting entails two phenomena:
